### 2015-05-22 doc/thesis/typedlua.tex
Improving the text about operation modes and evaluation categories

Even though these operation modes affect function calls, they do not
affect other multiple assignments.
For instance,
\begin{verbatim}
    local x:integer, y:integer = 1, 2, 3
\end{verbatim}
compiles without any warnings in both operation modes.
This means that the left-hand side of this local declaration has the
tuple type \texttt{(integer, integer, value*)}, while the expression
list in the right-hand side of this local declaration has the
tuple type \texttt{(1, 2, 3, nil*)} in both operation modes.

As another example,
\begin{verbatim}
    local x:integer, y:integer = sum(2, 2)
\end{verbatim}
compiles with warnings in both operation modes.
This means that the left-hand side of this local declaration has
the tuple type \texttt{(integer, integer, value*)}, while the
expression list in the right-hand side of this local declaration
has the tuple type \texttt{(integer, nil*)} in both operation modes.
The type \texttt{nil*} generates a value of type \texttt{nil}
that is not consistent with \texttt{integer}, the type of \texttt{y}.

@@

As an example,
\begin{verbatim}
    local x:integer, y:integer, z:string = m(), m()
\end{verbatim}
compiles without warnings in both operation modes.
This happens because in the right-hand side of the multiple
assignment, only the first value produced by the first call
to \texttt{m} gets used, so the type of the right-hand side
is \texttt{(integer, integer, string, nil*)}, which is
consistent with the type of the left-hand side
\texttt{(integer, integer, string, value*)}.

@@

We can also type variadic functions.
For instance,
\begin{verbatim}
    local function v (...:string):(string*)
      return ...
    end
\end{verbatim}
has type \texttt{(string*) -> (string*)} in both operation modes.
The call \texttt{v()} compiles without any warnings in the default and
strict modes because \texttt{(value*)} and \texttt{()} are consistent
with \texttt{(string*)}.
The call \texttt{v("hello", "world")} compiles without any warnings
in the default and strict modes because \texttt{("hello", "world", nil*)}
and \texttt{("hello", "world")} are consistent with \texttt{(string*)}.
The call \texttt{v(...)} compiles without any warnings in both
operation modes because the type of the argument list is \texttt{(string*)},
assuming that the vararg expression (\texttt{...}) has type \texttt{string}.

@@

The type of \texttt{message} is
\texttt{(string, string|nil, value*) -> (string, nil*)}
in the default mode, and
\texttt{(string, string|nil) -> (string)}
in the strict mode.
The call \texttt{message("Typed Lua")} compiles without
any warnings in both modes, because the argument list
\texttt{("Typed Lua", nil*)} is consistent with the input type
\texttt{(string, string|nil, value*)} in the default mode,
and it is also consistent with the input type
\texttt{(string, string|nil)} in the strict mode.
The compiler includes \texttt{nil*} in the tail of
an argument list in both modes, as it is necessary
to replace any optional parameters that may appear in
a function declaration.
In Section \ref{sec:assignment} we will formalize this behavior.

@@

So far, Typed Lua replaces projection types with the union of the
corresponding component of each tuple, when it infers that a variable
has a projection type.
In our example, the variables \texttt{q} and \texttt{r} get projection
types that map to the first and second components of the union of
tuple types \texttt{(integer, integer, nil*)|(nil, string, nil*)}.
This means that variables \texttt{q} and \texttt{r} have types
\texttt{integer|nil} and \texttt{integer|string}, respectively.

If a variable with a projection type appears in a \texttt{type} predicate,
it discriminates against all tuples in the projected union.
In our example, the projected union has type
\texttt{(integer, integer, nil*)|(nil, string, nil*)} outside of the
\texttt{if} statement,
but \texttt{(integer, integer, nil*)} inside the \texttt{then} block,
and \texttt{(nil, string, nil*)} inside the \texttt{else} block.
Thus, variable \texttt{q} has type \texttt{integer|nil} and variable
\texttt{r} has type \texttt{integer|string} outside of the
\texttt{if} statement;
but variable \texttt{q} has type \texttt{integer} and variable \texttt{r} also
has type \texttt{integer} inside the \texttt{then} block;
and variable \texttt{q} has type \texttt{nil} while variable \texttt{r}
has type \texttt{string} inside the \texttt{else} block.

@@

This example means that we cannot write a function that is
guaranteed to return a number when we pass a number and
guaranteed to return a string when we pass a string.
Even though we could use intersection types to express the type
of this function as
\texttt{(number) -> (number) \string& (string) -> (string)},
intersection types require more sophisticated flow typing to
check whether a function has this type,
and we still need to work on this problem.

>>>>

Even though these operation modes affect function calls, they do not
affect other multiple assignments.
This means that both operation modes discard extra arguments in
multiple assignments and they also use \texttt{nil} to replace missing
arguments in multiple assignments.
We made this design decision to avoid being to restrictive, that is,
we did not want to give unnecessary warnings to programmers,
as discarded values do not have any implications during run-time.

As an example,
\begin{verbatim}
    local x:integer, y:integer = 1, 2, 3
\end{verbatim}
compiles without any compile-time warnings in both operation modes.
This means that the left-hand side of this local declaration has the
tuple type \texttt{(integer, integer, value*)}, while the expression
list in the right-hand side of this local declaration has the tuple
type \texttt{(1, 2, 3, nil*)} in both operation modes.
The literal type \texttt{3} is consistent with \texttt{value},
and the type \texttt{nil*} generates a value of type \texttt{nil}
that is also consistent with \texttt{value}.

As another example,
\begin{verbatim}
    local x:integer, y:integer = sum(2, 2)
\end{verbatim}
compiles with one compile-time warning in both of the operation modes.
This means that the left-hand side of this local declaration has
the tuple type \texttt{(integer, integer, value*)}, while the
expression list in the right-hand side of this local declaration
has the tuple type \texttt{(integer, nil*)} in both operation modes.
The type \texttt{nil*} generates a value of type \texttt{nil}
that is not consistent with \texttt{integer}, the type of \texttt{y}.

@@

As an example,
\begin{verbatim}
    local x:integer, y:integer, z:string = m(), m()
\end{verbatim}
compiles without compile-time warnings in both operation modes.
This happens because in the right-hand side of the multiple
assignment, only the first value produced by the first call
to \texttt{m} gets used, so the type of the right-hand side
is \texttt{(integer, integer, string, nil*)}, which is
consistent with the type of the left-hand side
\texttt{(integer, integer, string, value*)}.

@@

We can also type variadic functions.
For instance,
\begin{verbatim}
    local function v (...:string):(string*)
      return ...
    end
\end{verbatim}
has type \texttt{(string*) -> (string*)} in both operation modes.
The function call \texttt{v()} compiles without any compile-time warnings in
the default and strict modes because \texttt{(value*)} and \texttt{()}
are consistent with \texttt{(string*)}.
The call \texttt{v("hello", "world")} compiles without any warnings
in the default and strict modes because \texttt{("hello", "world", nil*)}
and \texttt{("hello", "world")} are consistent with \texttt{(string*)}.
Calling \texttt{v(...)} compiles without any warnings in both
operation modes because the type of the argument list is \texttt{(string*)},
assuming that the vararg expression (\texttt{...}) has type \texttt{string}.

@@

The type of the function \texttt{message} is
\texttt{(string, string|nil, value*) -> (string, nil*)}
in the default mode, and
\texttt{(string, string|nil) -> (string)}
in the strict mode.
The function call \texttt{message("Typed Lua")} compiles without
any compile-time warnings in both modes, because the argument list
\texttt{("Typed Lua")} has type \texttt{("Typed Lua", nil*)} that
is consistent with the input type \texttt{(string, string|nil, value*)}
in the default mode, and that is also consistent with the input type
\texttt{(string, string|nil)} in the strict mode.
The compiler includes \texttt{nil*} in the tail of
an argument list in both modes, as it is necessary
to replace any optional parameters that may appear in
a function declaration.
In Section \ref{sec:assignment} we will formalize this behavior.

@@

So far, Typed Lua replaces projection types with the union of the
corresponding component of each tuple, when it infers that a local variable
has a projection type.
In our example, the variables \texttt{q} and \texttt{r} get projection
types that map to the first and second components of the union of
tuple types \texttt{(integer, integer, nil*)|(nil, string, nil*)}.
This means that variables \texttt{q} and \texttt{r} have types
\texttt{integer|nil} and \texttt{integer|string}, respectively.

If a variable with a projection type appears in a \texttt{type} predicate,
it discriminates all tuples in the projected union.
In our example, the projected union has type
\texttt{(integer, integer, nil*)|(nil, string, nil*)} outside of the
\texttt{if} statement,
but \texttt{(integer, integer, nil*)} inside the \texttt{then} block,
and \texttt{(nil, string, nil*)} inside the \texttt{else} block.
Thus, variable \texttt{q} has type \texttt{integer|nil} and variable
\texttt{r} has type \texttt{integer|string} outside of the
\texttt{if} statement;
but variable \texttt{q} has type \texttt{integer} and variable \texttt{r} also
has type \texttt{integer} inside the \texttt{then} block;
and variable \texttt{q} has type \texttt{nil} while variable \texttt{r}
has type \texttt{string} inside the \texttt{else} block.

@@

This example means that we cannot write a function that is
guaranteed to return a number when we pass a number and
guaranteed to return a string when we pass a string.
Even though we could have used an intersection type
\texttt{(number) -> (number) \string& (string) -> (string)}
to express the type of this function,
intersection types require more sophisticated flow typing to
check whether a function has this type,
and we still need to work on this problem.
